Mechanised Modal Model Theory

### 2019-11-17 p-modalmodels/paper.tex
paper 1st draft before 2nd meeting

\documentclass{llncs}

\begin{document}

@@

\begin{abstract}
\end{abstract}

@@

\section{Introduction}[1.5p]

@@

\section{Syntax, Semantics and the Standard Translation}[1.5p]

\section{Basic Results}[2p]

\section{Mechanizing Ultrafilters and Ultraproducts}[1p]

\section{Bisimilarity and Ultrafilter Extensions}[2p]

\section{Saturation of Ultraproduct Models}[3p]

\section{Van Benthem's Characterization Theorem}[3p]

>>>>

\documentclass{llncs}
\usepackage{holtexbasic}
\usepackage{stmaryrd}
\newtheorem{thm}{Theorem}[chapter]
\newtheorem{defn}{Definition}[chapter]
\newtheorem{lm}[thm]{Lemma}
\newtheorem{coro}[thm]{Corollary}
\newtheorem{prop}[thm]{Proposition}
\newenvironment{holmath}{\begin{displaymath}\begin{array}{l}}{\end{array}\end{displaymath}\ignorespacesafterend}
\renewcommand{\HOLConst}[1]{{\textsf{\upshape\small #1}}}
\renewcommand{\HOLinline}[1]{\ensuremath{#1}}
\renewcommand{\HOLFieldName}[1]{\HOLConst{#1}}
\renewcommand{\HOLTokenLeftrec}{\ensuremath{\langle\!\langle}}
\renewcommand{\HOLTokenRightrec}{\ensuremath{\rangle\!\rangle}}
\newcommand{\holthmenv}[1]{\begin{array}[t]{l}#1\end{array}}
\renewcommand{\HOLKeyword}[1]{\texttt{\upshape #1}}
\renewcommand{\HOLTokenBar}{\mbox{$|$}}
\begin{document}

@@

%\begin{abstract}
%\end{abstract}

@@

\section{Introduction}[1p]

@@

\section{Syntax, Semantics and the Standard Translation}[1.5p]
In our formalization, we only consider the basic modal language, in which the only primitive modal operator is the `$\Diamond$'. For a type $\alpha$, an $\alpha$-modal formula is either of form \HOLinline{\HOLConst{VAR}\;\HOLFreeVar{p}}, where $p$ is of type $\alpha$, or a disjunction $\phi\lor \psi$ of two $\alpha$-modal formulas, or the falsity $\bot$, or a negation $\lnot \phi$ of an $\alpha$-modal formula $\phi$, or, finally, of the form $\Diamond \phi$ where $\phi$ is an $\alpha$-modal formula. We create a data type called `form' of the formulas of this modal language. 
\begin{defn}
{\upshape\cite[Definition 1.9]{Blackburn}}
\begin{holmath}
  \ensuremath{\alpha}\;\HOLTyOp{form}\;=\;\HOLConst{VAR}\;\ensuremath{\alpha}\;\HOLTokenBar{}\;\HOLConst{DISJ}\;(\ensuremath{\alpha}\;\HOLTyOp{form})\;(\ensuremath{\alpha}\;\HOLTyOp{form})\;\HOLTokenBar{}\;\HOLSymConst{\ensuremath{\bot}}\;\HOLTokenBar{}\;(\HOLSymConst{\HOLTokenNeg{}})\;(\ensuremath{\alpha}\;\HOLTyOp{form})\;\HOLTokenBar{}\;\HOLConst{\ensuremath{\Diamond}}\;(\ensuremath{\alpha}\;\HOLTyOp{form})
\end{holmath}
\end{defn}

A model where these formulas can be interpreted consists of a \emph{frame} and a \emph{valuation}, where \HOLinline{\HOLConst{frame}} is a set with a relation on it. 
\begin{defn}
{\upshape\cite[Definition 1.19]{Blackburn}}
\begin{holmath}
  \ensuremath{\beta}\;\HOLTyOp{frame} = \HOLTokenLeftrec{}\;\HOLFieldName{world}\;{:}\;\ensuremath{\beta}\;\ensuremath{\rightarrow}\;\HOLTyOp{bool};\;\HOLFieldName{rel}\;{:}\;\ensuremath{\beta}\;\ensuremath{\rightarrow}\;\ensuremath{\beta}\;\ensuremath{\rightarrow}\;\HOLTyOp{bool}\;\HOLTokenRightrec{}\\
  (\ensuremath{\alpha},\;\ensuremath{\beta})\;\HOLTyOp{model} = \HOLTokenLeftrec{}\;\HOLFieldName{frame}\;{:}\;\ensuremath{\beta}\;\HOLTyOp{frame};\;\HOLFieldName{valt}\;{:}\;\ensuremath{\alpha}\;\ensuremath{\rightarrow}\;\ensuremath{\beta}\;\ensuremath{\rightarrow}\;\HOLTyOp{bool}\;\HOLTokenRightrec{}
\end{holmath}
\end{defn}
In the rest of the paper, the field \HOLinline{\HOLFreeVar{\ensuremath{\mathfrak{M}}}}$.\sf valt$ of a model \HOLinline{\HOLFreeVar{\ensuremath{\mathfrak{M}}}} will be called the \emph{valuation}, and \HOLinline{\HOLFreeVar{\ensuremath{\mathfrak{M}}}\HOLSymConst{\ensuremath{\sp{W}}}}, \HOLinline{\HOLFreeVar{\ensuremath{\mathfrak{M}}}\HOLSymConst{\ensuremath{\sp{R}}}} and \HOLinline{\HOLFreeVar{\ensuremath{\mathfrak{M}}}\HOLSymConst{\ensuremath{\sp{V}}}} are used to denote the world set, the relation, and the valuation of \HOLinline{\HOLFreeVar{\ensuremath{\mathfrak{M}}}} respectively.  When we say a $(\HOLinline{\ensuremath{\alpha}},\HOLinline{\ensuremath{\beta}})$-model, we mean a model for $\alpha$-formulas with a $\beta$-set as its underlying set. The interpretation of $\alpha$-modal formulas on an $(\HOLinline{\ensuremath{\alpha}},\HOLinline{\ensuremath{\beta}})$-model is given by the predicate \emph{satisfaction}. We read `\HOLinline{\HOLFreeVar{\ensuremath{\mathfrak{M}}},\HOLFreeVar{w}\ensuremath{\Vdash}\;\HOLFreeVar{\ensuremath{\phi}}}' as `$\phi$ is satisfied at the world $w$ in \HOLinline{\HOLFreeVar{\ensuremath{\mathfrak{M}}}}'.
\begin{defn}
{\upshape\cite[Definition 1.20]{Blackburn}}
\begin{holmath}
\begin{array}{rcl}
  \HOLFreeVar{\ensuremath{\mathfrak{M}}},\HOLFreeVar{w}\ensuremath{\Vdash}\;\HOLConst{VAR}\;\HOLFreeVar{p}\;&\HOLTokenDefEquality{}&\;\HOLFreeVar{w}\;\HOLSymConst{\HOLTokenIn{}}\;\HOLFreeVar{\ensuremath{\mathfrak{M}}}\HOLSymConst{\ensuremath{\sp{W}}}\;\HOLSymConst{\HOLTokenConj{}}\;\HOLFreeVar{w}\;\HOLSymConst{\HOLTokenIn{}}\;\HOLFreeVar{\ensuremath{\mathfrak{M}}}\HOLSymConst{\ensuremath{\sp{V}}}\;\HOLFreeVar{p}\\
\HOLFreeVar{\ensuremath{\mathfrak{M}}},\HOLFreeVar{w}\ensuremath{\Vdash}\;\HOLSymConst{\ensuremath{\bot}}\;&\HOLTokenDefEquality{}&\;\HOLFreeVar{w}\;\HOLSymConst{\HOLTokenIn{}}\;\HOLFreeVar{\ensuremath{\mathfrak{M}}}\HOLSymConst{\ensuremath{\sp{W}}}\;\HOLSymConst{\HOLTokenConj{}}\;\HOLConst{F}\\
\HOLFreeVar{\ensuremath{\mathfrak{M}}},\HOLFreeVar{w}\ensuremath{\Vdash}\;\HOLSymConst{\HOLTokenNeg{}}\HOLFreeVar{\ensuremath{\phi}}\;&\HOLTokenDefEquality{}&\;\HOLFreeVar{w}\;\HOLSymConst{\HOLTokenIn{}}\;\HOLFreeVar{\ensuremath{\mathfrak{M}}}\HOLSymConst{\ensuremath{\sp{W}}}\;\HOLSymConst{\HOLTokenConj{}}\;\HOLFreeVar{\ensuremath{\mathfrak{M}}},\HOLFreeVar{w}\ensuremath{\not\Vdash}\;\HOLFreeVar{\ensuremath{\phi}}\\
\HOLFreeVar{\ensuremath{\mathfrak{M}}},\HOLFreeVar{w}\ensuremath{\Vdash}\;(\HOLFreeVar{\ensuremath{\phi\sb{1}}}\HOLSymConst{\ensuremath{\lor}}\HOLFreeVar{\ensuremath{\phi\sb{2}}})\;&\HOLTokenDefEquality{}&\;\HOLFreeVar{\ensuremath{\mathfrak{M}}},\HOLFreeVar{w}\ensuremath{\Vdash}\;\HOLFreeVar{\ensuremath{\phi\sb{1}}}\;\HOLSymConst{\HOLTokenDisj{}}\;\HOLFreeVar{\ensuremath{\mathfrak{M}}},\HOLFreeVar{w}\ensuremath{\Vdash}\;\HOLFreeVar{\ensuremath{\phi\sb{2}}}\\
\HOLFreeVar{\ensuremath{\mathfrak{M}}},\HOLFreeVar{w}\ensuremath{\Vdash}\;\HOLSymConst{\ensuremath{\Diamond}}\HOLFreeVar{\ensuremath{\phi}}\;&\HOLTokenDefEquality{}&\;\HOLFreeVar{w}\;\HOLSymConst{\HOLTokenIn{}}\;\HOLFreeVar{\ensuremath{\mathfrak{M}}}\HOLSymConst{\ensuremath{\sp{W}}}\;\HOLSymConst{\HOLTokenConj{}}\;\HOLSymConst{\HOLTokenExists{}}\HOLBoundVar{v}.\;\HOLFreeVar{\ensuremath{\mathfrak{M}}}\HOLSymConst{\ensuremath{\sp{R}}}\;\HOLFreeVar{w}\;\HOLBoundVar{v}\;\HOLSymConst{\HOLTokenConj{}}\;\HOLBoundVar{v}\;\HOLSymConst{\HOLTokenIn{}}\;\HOLFreeVar{\ensuremath{\mathfrak{M}}}\HOLSymConst{\ensuremath{\sp{W}}}\;\HOLSymConst{\HOLTokenConj{}}\;\HOLFreeVar{\ensuremath{\mathfrak{M}}},\HOLBoundVar{v}\ensuremath{\Vdash}\;\HOLFreeVar{\ensuremath{\phi}}
\end{array}
\end{holmath}
\end{defn}
If \HOLinline{\HOLFreeVar{\ensuremath{\phi\sb{1}}}}, \HOLinline{\HOLFreeVar{\ensuremath{\phi\sb{2}}}} are \HOLinline{\HOLFreeVar{\ensuremath{\alpha}}}-formulas, we say they are \emph{equivalent} if for every $(\alpha,\beta)$-model \HOLinline{\HOLFreeVar{\ensuremath{\mathfrak{M}}}} and every world \HOLinline{\HOLFreeVar{w}} in it, we have \HOLinline{\HOLFreeVar{\ensuremath{\mathfrak{M}}},\HOLFreeVar{w}\ensuremath{\Vdash}\;\HOLFreeVar{\ensuremath{\phi\sb{1}}}\;\HOLSymConst{\HOLTokenEquiv{}}\;\HOLFreeVar{\ensuremath{\mathfrak{M}}},\HOLFreeVar{w}\ensuremath{\Vdash}\;\HOLFreeVar{\ensuremath{\phi\sb{2}}}}.
\begin{defn}[Equivalence]
\begin{holmath}
  (\HOLFreeVar{\ensuremath{\phi\sb{1}}} :\ensuremath{\alpha}\;\HOLTyOp{form})\equiv\sb{({:}\ensuremath{\beta})}(\HOLFreeVar{\ensuremath{\phi\sb{2}}} :\ensuremath{\alpha}\;\HOLTyOp{form})\;\HOLTokenDefEquality{}\\
\;\;\HOLSymConst{\HOLTokenForall{}}(\HOLBoundVar{\ensuremath{\mathfrak{M}}} :(\ensuremath{\alpha},\;\ensuremath{\beta})\;\HOLTyOp{model})\;(\HOLBoundVar{w} :\ensuremath{\beta}).\;\HOLBoundVar{\ensuremath{\mathfrak{M}}},\HOLBoundVar{w}\ensuremath{\Vdash}\;\HOLFreeVar{\ensuremath{\phi\sb{1}}}\;\HOLSymConst{\HOLTokenEquiv{}}\;\HOLBoundVar{\ensuremath{\mathfrak{M}}},\HOLBoundVar{w}\ensuremath{\Vdash}\;\HOLFreeVar{\ensuremath{\phi\sb{2}}}\\[3mm]
\end{holmath}
\end{defn}
We cannot omit the type parameter \HOLinline{({:}\ensuremath{\beta})} in the definition, since then there will be a type, namely the type of the underlying set of the models we are talking about, that only appears on the right-hand side but not on the left-hand side of the definition, which is not allowed in HOL. Also, we are not allowed to quantify over types, so it is also impossible to define the equivalence to be \HOLinline{\HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{\ensuremath{\mu}}.\;\HOLFreeVar{\ensuremath{\phi\sb{1}}}\equiv\sb{\HOLBoundVar{\ensuremath{\mu}}}\HOLFreeVar{\ensuremath{\phi\sb{2}}}}, where $\mu$ denotes a type. Therefore, this definition is not encoding the equivalence in mathematical sense precisely: when we mention equivalence of formulas in usual mathematical language, we are implicitly referring to the class of all models, but the constraint here bans us from talking about all models of all possible types at once.

A modal formula can be translated into a first-order formula via the \emph{standard translation}. A translated formula has at most one free variable. For a modal formula \HOLinline{\HOLFreeVar{\ensuremath{\phi}}}, \HOLinline{{\sf ST}\sb{\HOLFreeVar{x}}\;\HOLFreeVar{\ensuremath{\phi}}} is the standard translation of \HOLinline{\HOLFreeVar{\ensuremath{\phi}}} using \HOLinline{\HOLFreeVar{x}} as the only free variable.
\begin{defn}
{\upshape\cite[Definition 2.45 (Standard Translation)]{Blackburn}}
\begin{holmath}
\begin{array}{rcl}
  {\sf ST}\sb{\HOLFreeVar{x}}\;(\HOLConst{VAR}\;\HOLFreeVar{p})\;&\HOLTokenDefEquality{}&\;\HOLConst{\ensuremath{{}\sp{\mathsf{f}}\mathsf{P}}}\;\HOLFreeVar{p}\;(\HOLConst{\ensuremath{{}^{\sf f}{\sf VAR}}}\;\HOLFreeVar{x})\\
{\sf ST}\sb{\HOLFreeVar{x}}\;\HOLSymConst{\ensuremath{\bot}}\;&\HOLTokenDefEquality{}&\;\HOLConst{\ensuremath{{}^{\sf f}\bot}}\\
{\sf ST}\sb{\HOLFreeVar{x}}\;(\HOLSymConst{\HOLTokenNeg{}}\HOLFreeVar{\ensuremath{\phi}})\;&\HOLTokenDefEquality{}&\;\HOLConst{\ensuremath{{^{\sf f}}{\lnot}}}\;({\sf ST}\sb{\HOLFreeVar{x}}\;\HOLFreeVar{\ensuremath{\phi}})\\
{\sf ST}\sb{\HOLFreeVar{x}}\;(\HOLFreeVar{\ensuremath{\phi}}\HOLSymConst{\ensuremath{\lor}}\HOLFreeVar{\ensuremath{\psi}})\;&\HOLTokenDefEquality{}&\;{\sf ST}\sb{\HOLFreeVar{x}}\;\HOLFreeVar{\ensuremath{\phi}}\HOLSymConst{\ensuremath{\ {}^{\sf f} {\lor} \ }}{\sf ST}\sb{\HOLFreeVar{x}}\;\HOLFreeVar{\ensuremath{\psi}}\\
{\sf ST}\sb{\HOLFreeVar{x}}\;(\HOLSymConst{\ensuremath{\Diamond}}\HOLFreeVar{\ensuremath{\phi}})\;&\HOLTokenDefEquality{}&\;\HOLConst{\ensuremath{ {}^{\sf f}{\exists}}}\;(\HOLFreeVar{x}\;\HOLSymConst{\ensuremath{+}}\;\HOLNumLit{1})\;(\HOLConst{\ensuremath{{}^{\sf f}{\sf R}}}\;(\HOLConst{\ensuremath{{}^{\sf f}{\sf VAR}}}\;\HOLFreeVar{x})\;(\HOLConst{\ensuremath{{}^{\sf f}{\sf VAR}}}\;(\HOLFreeVar{x}\;\HOLSymConst{\ensuremath{+}}\;\HOLNumLit{1}))\HOLSymConst{\ensuremath{\ {}^{\sf f}{\land} \ }}{\sf ST}\sb{\HOLFreeVar{x}\;\HOLSymConst{\ensuremath{+}}\;\HOLNumLit{1}}\;\HOLFreeVar{\ensuremath{\phi}})
\end{array}
\end{holmath}
\end{defn}
According to the semantic interpretation of the `\HOLinline{\HOLSymConst{\ensuremath{\Diamond}}\HOLFreeVar{\ensuremath{\phi}}}', we translate \HOLinline{\HOLSymConst{\ensuremath{\Diamond}}\HOLFreeVar{\ensuremath{\phi}}} into the existential formula. To make sure that we use a fresh variable symbol that is not the same as the variable $x$ which is marking the current state, we use $x+1$ as our new variable symbol. The standard translation gives a first-order reformulation of satisfaction of modal formulas:
\begin{prop}\label{2.47}
{\upshape\cite[Theorem 2.47 (i)]{Blackburn}}
\begin{holmath}
  \ensuremath{\HOLTokenTurnstile}\HOLFreeVar{\ensuremath{\mathfrak{M}}},\HOLFreeVar{w}\ensuremath{\Vdash}\;\HOLFreeVar{\ensuremath{\phi}}\;\HOLSymConst{\HOLTokenEquiv{}}\;\HOLConst{mm2folm}\;\HOLFreeVar{\ensuremath{\mathfrak{M}}},(\HOLTokenLambda{}\HOLBoundVar{n}.\;\HOLFreeVar{w})\ensuremath{\vDash}\;{\sf ST}\sb{\HOLFreeVar{x}}\;\HOLFreeVar{\ensuremath{\phi}}
\end{holmath}
\end{prop} 
Here \HOLinline{\HOLConst{mm2folm}} is the function that turns a modal model into a first-order model, defined as:
\begin{holmath}
  \HOLConst{mm2folm}\;\HOLFreeVar{\ensuremath{\mathfrak{M}}}\;\HOLTokenDefEquality{}\\
\;\;\HOLTokenLeftrec{}\HOLFieldName{Dom}\;:=\;\HOLFreeVar{\ensuremath{\mathfrak{M}}}\HOLSymConst{\ensuremath{\sp{W}}};\;\HOLFieldName{Fun}\;:=\;(\HOLTokenLambda{}\HOLBoundVar{n}\;\HOLBoundVar{l}.\;\HOLConst{CHOICE}\;\HOLFreeVar{\ensuremath{\mathfrak{M}}}\HOLSymConst{\ensuremath{\sp{W}}});\\
\;\;\HOLFieldName{Pred}\;:=\\
\;\;\;\;(\HOLTokenLambda{}\HOLBoundVar{p}\;\HOLBoundVar{zs}.\\
\;\;\;\;\;\;\;\;\;\HOLKeyword{case}\;\HOLBoundVar{zs}\;\HOLKeyword{of}\\
\;\;\;\;\;\;\;\;\;\;\;[]\;\HOLTokenImp{}\;\HOLConst{F}\\
\;\;\;\;\;\;\;\;\;\HOLTokenBar{}\;[\HOLBoundVar{w}\sb{\mathrm{1}}]\;\HOLTokenImp{}\;\HOLBoundVar{w}\sb{\mathrm{1}}\;\HOLSymConst{\HOLTokenIn{}}\;\HOLFreeVar{\ensuremath{\mathfrak{M}}}\HOLSymConst{\ensuremath{\sp{W}}}\;\HOLSymConst{\HOLTokenConj{}}\;\HOLFreeVar{\ensuremath{\mathfrak{M}}}\HOLSymConst{\ensuremath{\sp{V}}}\;\HOLBoundVar{p}\;\HOLBoundVar{w}\sb{\mathrm{1}}\\
\;\;\;\;\;\;\;\;\;\HOLTokenBar{}\;[\HOLBoundVar{w}\sb{\mathrm{1}};\;\HOLBoundVar{w}\sb{\mathrm{2}}]\;\HOLTokenImp{}\;\HOLBoundVar{p}\;\HOLSymConst{=}\;\HOLNumLit{0}\;\HOLSymConst{\HOLTokenConj{}}\;\HOLFreeVar{\ensuremath{\mathfrak{M}}}\HOLSymConst{\ensuremath{\sp{R}}}\;\HOLBoundVar{w}\sb{\mathrm{1}}\;\HOLBoundVar{w}\sb{\mathrm{2}}\;\HOLSymConst{\HOLTokenConj{}}\;\HOLBoundVar{w}\sb{\mathrm{1}}\;\HOLSymConst{\HOLTokenIn{}}\;\HOLFreeVar{\ensuremath{\mathfrak{M}}}\HOLSymConst{\ensuremath{\sp{W}}}\;\HOLSymConst{\HOLTokenConj{}}\;\HOLBoundVar{w}\sb{\mathrm{2}}\;\HOLSymConst{\HOLTokenIn{}}\;\HOLFreeVar{\ensuremath{\mathfrak{M}}}\HOLSymConst{\ensuremath{\sp{W}}}\\
\;\;\;\;\;\;\;\;\;\HOLTokenBar{}\;\HOLBoundVar{w}\sb{\mathrm{1}}\HOLSymConst{::}\HOLBoundVar{w}\sb{\mathrm{2}}\HOLSymConst{::}\HOLBoundVar{w}\sb{\mathrm{3}}\HOLSymConst{::}\HOLBoundVar{w}\sb{\mathrm{4}}\;\HOLTokenImp{}\;\HOLConst{F})\HOLTokenRightrec{}
\end{holmath}

\section{Basic Results}%[2p]
We discuss some highlights of formalized results in section 2.1-2.3 in \cite{Blackburn} below.
\subsection{Tree-like property}
A tree-like model is a model whose underlying frame is a tree. If $H$ is a tree with root $r$, we write \HOLinline{\HOLConst{tree}\;\HOLFreeVar{H}\;\HOLFreeVar{r}} in HOL:
\begin{defn}
{\upshape\cite[Definition 1.7]{Blackburn}}
\begin{holmath}
   \HOLConst{tree}\;\HOLFreeVar{H}\;\HOLFreeVar{r}\;\HOLTokenDefEquality{}\\
\;\;\HOLFreeVar{r}\;\HOLSymConst{\HOLTokenIn{}}\;\HOLFreeVar{H}.\HOLFieldName{world}\;\HOLSymConst{\HOLTokenConj{}}\;(\HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{w}.\;\HOLBoundVar{w}\;\HOLSymConst{\HOLTokenIn{}}\;\HOLFreeVar{H}.\HOLFieldName{world}\;\HOLSymConst{\HOLTokenImp{}}\;{\HOLFreeVar{H}.\HOLFieldName{rel}}\ensuremath{\mid}\sb{\HOLFreeVar{H}.\HOLFieldName{world}}\HOLSymConst{\HOLTokenSupStar{}}\;\HOLFreeVar{r}\;\HOLBoundVar{w})\;\HOLSymConst{\HOLTokenConj{}}\\
\;\;(\HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{w}.\;\HOLBoundVar{w}\;\HOLSymConst{\HOLTokenIn{}}\;\HOLFreeVar{H}.\HOLFieldName{world}\;\HOLSymConst{\HOLTokenImp{}}\;\HOLSymConst{\HOLTokenNeg{}}\HOLFreeVar{H}.\HOLFieldName{rel}\;\HOLBoundVar{w}\;\HOLFreeVar{r})\;\HOLSymConst{\HOLTokenConj{}}\\
\;\;\HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{w}.\;\HOLBoundVar{w}\;\HOLSymConst{\HOLTokenIn{}}\;\HOLFreeVar{H}.\HOLFieldName{world}\;\HOLSymConst{\HOLTokenConj{}}\;\HOLBoundVar{w}\;\HOLSymConst{\HOLTokenNotEqual{}}\;\HOLFreeVar{r}\;\HOLSymConst{\HOLTokenImp{}}\;\HOLSymConst{\HOLTokenUnique{}}\HOLBoundVar{w}\sb{\mathrm{0}}.\;\HOLBoundVar{w}\sb{\mathrm{0}}\;\HOLSymConst{\HOLTokenIn{}}\;\HOLFreeVar{H}.\HOLFieldName{world}\;\HOLSymConst{\HOLTokenConj{}}\;\HOLFreeVar{H}.\HOLFieldName{rel}\;\HOLBoundVar{w}\sb{\mathrm{0}}\;\HOLBoundVar{w}
\end{holmath}
\end{defn}
The tree-like property says an satisfiable modal formula can be satisfied in a tree-like model. It is proved in HOL as:
\begin{prop}
{\upshape\cite[Proposition 2.15]{Blackburn}}\label{2.15}
\begin{holmath}
  \ensuremath{\HOLTokenTurnstile}(\HOLFreeVar{\ensuremath{\mathfrak{M}}}\sb{\mathrm{1}} :(\ensuremath{\alpha},\;\ensuremath{\beta})\;\HOLTyOp{model}),(\HOLFreeVar{w} :\ensuremath{\beta})\ensuremath{\Vdash}\;(\HOLFreeVar{\ensuremath{\phi}} :\ensuremath{\alpha}\;\HOLTyOp{form})\;\HOLSymConst{\HOLTokenImp{}}\\
\;\;\;\;\;\;\;\HOLSymConst{\HOLTokenExists{}}(\HOLBoundVar{\ensuremath{\mathfrak{M}}} :(\ensuremath{\alpha},\;\ensuremath{\beta}\;\HOLTyOp{list})\;\HOLTyOp{model})\;(\HOLBoundVar{r} :\ensuremath{\beta}\;\HOLTyOp{list}).\;\HOLConst{tree}\;\HOLBoundVar{\ensuremath{\mathfrak{M}}}.\HOLFieldName{frame}\;\HOLBoundVar{r}\;\HOLSymConst{\HOLTokenConj{}}\;\HOLBoundVar{\ensuremath{\mathfrak{M}}},\HOLBoundVar{r}\ensuremath{\Vdash}\;\HOLFreeVar{\ensuremath{\phi}}
\end{holmath}
\end{prop}
The world set of the tree like model constructed from \HOLinline{\HOLFreeVar{\ensuremath{\mathfrak{M}}}} according to the standard proof of this theorem is a set of lists of worlds in \HOLinline{\HOLFreeVar{\ensuremath{\mathfrak{M}}}\sb{\mathrm{1}}}, that is why passing to a tree-like model does not preserve the type.

\subsection{Bisimulation}
The definition of bisimulation in HOL is just a direct translation of its mathematical definition: 
\begin{defn}
{\upshape\cite[Definition 2.16 (Bisimulations)]{Blackburn}}
\begin{holmath}
  \HOLFreeVar{\ensuremath{\mathfrak{M}}}\sb{\mathrm{1}}\overset{\HOLFreeVar{Z}}\leftrightarroweq\HOLFreeVar{\ensuremath{\mathfrak{M}}}\sb{\mathrm{2}}\;\HOLTokenDefEquality{}\\
\;\;\HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{w}\sb{\mathrm{1}}\;\HOLBoundVar{w}\sb{\mathrm{2}}.\\
\;\;\;\;\;\;\HOLBoundVar{w}\sb{\mathrm{1}}\;\HOLSymConst{\HOLTokenIn{}}\;\HOLFreeVar{\ensuremath{\mathfrak{M}}}\sb{\mathrm{1}}\HOLSymConst{\ensuremath{\sp{W}}}\;\HOLSymConst{\HOLTokenConj{}}\;\HOLBoundVar{w}\sb{\mathrm{2}}\;\HOLSymConst{\HOLTokenIn{}}\;\HOLFreeVar{\ensuremath{\mathfrak{M}}}\sb{\mathrm{2}}\HOLSymConst{\ensuremath{\sp{W}}}\;\HOLSymConst{\HOLTokenConj{}}\;\HOLFreeVar{Z}\;\HOLBoundVar{w}\sb{\mathrm{1}}\;\HOLBoundVar{w}\sb{\mathrm{2}}\;\HOLSymConst{\HOLTokenImp{}}\\
\;\;\;\;\;\;\;\;(\HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{p}.\;\HOLFreeVar{\ensuremath{\mathfrak{M}}}\sb{\mathrm{1}},\HOLBoundVar{w}\sb{\mathrm{1}}\ensuremath{\Vdash}\;\HOLConst{VAR}\;\HOLBoundVar{p}\;\HOLSymConst{\HOLTokenEquiv{}}\;\HOLFreeVar{\ensuremath{\mathfrak{M}}}\sb{\mathrm{2}},\HOLBoundVar{w}\sb{\mathrm{2}}\ensuremath{\Vdash}\;\HOLConst{VAR}\;\HOLBoundVar{p})\;\HOLSymConst{\HOLTokenConj{}}\\
\;\;\;\;\;\;\;\;(\HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{v}\sb{\mathrm{1}}.\\
\;\;\;\;\;\;\;\;\;\;\;\;\;\HOLBoundVar{v}\sb{\mathrm{1}}\;\HOLSymConst{\HOLTokenIn{}}\;\HOLFreeVar{\ensuremath{\mathfrak{M}}}\sb{\mathrm{1}}\HOLSymConst{\ensuremath{\sp{W}}}\;\HOLSymConst{\HOLTokenConj{}}\;\HOLFreeVar{\ensuremath{\mathfrak{M}}}\sb{\mathrm{1}}\HOLSymConst{\ensuremath{\sp{R}}}\;\HOLBoundVar{w}\sb{\mathrm{1}}\;\HOLBoundVar{v}\sb{\mathrm{1}}\;\HOLSymConst{\HOLTokenImp{}}\\
\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\HOLSymConst{\HOLTokenExists{}}\HOLBoundVar{v}\sb{\mathrm{2}}.\;\HOLBoundVar{v}\sb{\mathrm{2}}\;\HOLSymConst{\HOLTokenIn{}}\;\HOLFreeVar{\ensuremath{\mathfrak{M}}}\sb{\mathrm{2}}\HOLSymConst{\ensuremath{\sp{W}}}\;\HOLSymConst{\HOLTokenConj{}}\;\HOLFreeVar{Z}\;\HOLBoundVar{v}\sb{\mathrm{1}}\;\HOLBoundVar{v}\sb{\mathrm{2}}\;\HOLSymConst{\HOLTokenConj{}}\;\HOLFreeVar{\ensuremath{\mathfrak{M}}}\sb{\mathrm{2}}\HOLSymConst{\ensuremath{\sp{R}}}\;\HOLBoundVar{w}\sb{\mathrm{2}}\;\HOLBoundVar{v}\sb{\mathrm{2}})\;\HOLSymConst{\HOLTokenConj{}}\\
\;\;\;\;\;\;\;\;\HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{v}\sb{\mathrm{2}}.\\
\;\;\;\;\;\;\;\;\;\;\;\;\HOLBoundVar{v}\sb{\mathrm{2}}\;\HOLSymConst{\HOLTokenIn{}}\;\HOLFreeVar{\ensuremath{\mathfrak{M}}}\sb{\mathrm{2}}\HOLSymConst{\ensuremath{\sp{W}}}\;\HOLSymConst{\HOLTokenConj{}}\;\HOLFreeVar{\ensuremath{\mathfrak{M}}}\sb{\mathrm{2}}\HOLSymConst{\ensuremath{\sp{R}}}\;\HOLBoundVar{w}\sb{\mathrm{2}}\;\HOLBoundVar{v}\sb{\mathrm{2}}\;\HOLSymConst{\HOLTokenImp{}}\\
\;\;\;\;\;\;\;\;\;\;\;\;\;\;\HOLSymConst{\HOLTokenExists{}}\HOLBoundVar{v}\sb{\mathrm{1}}.\;\HOLBoundVar{v}\sb{\mathrm{1}}\;\HOLSymConst{\HOLTokenIn{}}\;\HOLFreeVar{\ensuremath{\mathfrak{M}}}\sb{\mathrm{1}}\HOLSymConst{\ensuremath{\sp{W}}}\;\HOLSymConst{\HOLTokenConj{}}\;\HOLFreeVar{Z}\;\HOLBoundVar{v}\sb{\mathrm{1}}\;\HOLBoundVar{v}\sb{\mathrm{2}}\;\HOLSymConst{\HOLTokenConj{}}\;\HOLFreeVar{\ensuremath{\mathfrak{M}}}\sb{\mathrm{1}}\HOLSymConst{\ensuremath{\sp{R}}}\;\HOLBoundVar{w}\sb{\mathrm{1}}\;\HOLBoundVar{v}\sb{\mathrm{1}}
\end{holmath}
\end{defn}

It is trivial to prove by induction that bisimilar worlds are modal equivalent. As the most significant theorem on the basic theory of bisimulations, we proved the Hennessy-Milner theorem as, which says modal equivalence and bisimulation on \emph{image finite} models are the same thing. An image-finite model is a model where every world can only be related to finitely many worlds. The Hennessy-Milner Theorem in HOL looks like:
